\bfseries\LARGE\singlespacing{Estimating \emph{Mycobacterium} tuberculosis transmission in a South African clinic: Spatiotemporal model based on person movements}

### 2024-05-21 doc/paper/manuscript.tex
Updates from Overleaf

\title{\bfseries\LARGE\singlespacing{Estimating \emph{Mycobacterium} tuberculosis transmission in a South African clinic: Spatiotemporal model}}
% author list
\author[1]{Nicolas Banholzer}
\author[2]{Keren Middelkoop}
\author[2]{Juane Leukes}
\author[3]{Ernest Weingartner}
\author[1]{Remo Schmutz}
\author[1]{Kathrin Zürcher}
\author[1,4,5]{Matthias Egger}
\author[2]{Robin Wood}
\author[1*]{Lukas Fenner}

@@

%TC:newcounter abst Words in abstract
%TC:envir abstract [] abst
\begin{abstract}\normalfont
The risk of Mycobacterium tuberculosis (\emph{Mtb}) transmission can be high in crowded clinics. We developed a spatiotemporal model of airborne \emph{Mtb} transmission, based on the Wells-Riley equation. We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when masks were compulsory and ventilation increased. We matched person tracks with clinical records to identify the spatiotemporal location of infectious TB patients. We modeled the concentration of infectious doses (quanta) and estimated the individual risk of infection. Over five days, video sensors detected 1,438 clinic attendees. CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm); the quanta concentration was higher in the morning than in the afternoon, and highest in the waiting room. The estimated risk of infection per clinic attendee was 0.06\% (95\%-credible interval [CrI] 0.001\%$-$0.40\%). It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.28-fold (95\%-CrI 1.19$-$1.36) higher in scenarios without mask use and 1.74-fold (95\%-CrI 0.47$-$5.34) with pre-pandemic ventilation rates. Spatiotemporal modeling can identify high-risk areas and evaluate the impact of infection control measures in clinics.

@@

Based on a pilot study from 2019 in the same clinic\cite{Zurcher2022JID,Zurcher2020BMJ}, we collected environmental data (indoor CO$_2$ levels), clinical data (TB status), and video sensor data (person tracks) for five days during the COVID-19 pandemic in October/November 2021 (October 13, 15, 25, and November 4, 5) at a primary care clinic in Cape Town, South Africa. The clinic offers both TB and HIV services and other basic clinical services, Monday through Friday, from 8\,am to 4\,pm. It is located in a large settlement of formal and semi-formal housing where TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}. We defined three areas within the clinic (\Cref{fig:floor-plan}): the waiting room (10.55m $\times$ 5.50m $\times$ 3.00m), corridor (12.45m $\times$ 2.20m $\times$ 2.50m), and TB room (4.75m $\times$ 3.50m $\times$ 3.00m). The TB room was mainly used for the treatment of diagnosed TB patients but also for the screening of suspected patients with respiratory symptoms who were ultimately not diagnosed with TB.

\begin{figure}[!htpb]
    \centering
    \includegraphics{doc/clinic-schematic-annotated-view.pdf}
    \caption{\textbf{Schematic view of the clinic}. Patients typically enter the clinic through the entrance on the bottom left, register at the reception, then wait in the waiting room or along the corridor until they see a doctor in one of the care rooms or the TB room. They exit the clinic through one of the exits in the waiting room or at the end of the corridor. Coloured icons show the placement of the video sensors to track persons and the devices monitoring CO$_2$ levels.}
    \label{fig:floor-plan}
\end{figure}

@@

We used an anonymous personal tracking system using video sensors (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1-second intervals (\Cref{fig:floor-plan}). The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic. People in the clinic could contribute multiple records if they moved out of the range of a sensor or were not recognized as a person. To address this, we created an R Shiny tool (\supp~Fig.~\zref{fig:shiny-app}) to rejoin interrupted tracks as described in \supp~Section~\zref{sec:setting-and-data}. We also labeled the tracks to identify the movements of clinic staff. \supp~Fig.~\zref{fig:tracking-examples} shows some sample tracks.

@@

\begin{figure}[!htpb]
    \centering
    \includegraphics{doc/model-flow-chart.pdf}
    \caption{\textbf{Visual summary of the spatiotemporal modelling approach}. (1)~Clinical records and person tracks were linked based on their timestamps at the registration to determine the spatiotemporal location of diagnosed (infectious) TB patients. (2)~Both diagnosed and randomly selected undiagnosed TB patients among clinic attendees generate infectious quanta. (3)~Person-tracking and environmental data were combined to estimate the air change rate. (4)~Air change rate was related to the speed at which quanta diffuses indoors. (5)~The air change rate also determines quanta removal through outdoor air exchange, in addition to inactivation of \emph{Mtb} in airborne particles. (6)~Quanta generation, diffusion, and removal determine the spatiotemporal quanta concentration, from which the personal risk of infection is calculated using the Wells-Riley equation.}
    \label{fig:modeling-flow}
\end{figure}

@@

(1)~We matched clinical and person-tracking data to identify the spatiotemporal location of infectious TB patients within the clinic. We recorded the timestamp when a person was detected in the registration area and matched this with the registration time in the clinical records. We excluded tracks with a total duration of less than five minutes in the clinic and a duration of less than five seconds in the registration area from the matching. We allowed for a maximum delay of 15\,min between the track in the registration area and the entry in the clinical database. All diagnosed TB patients could be matched with person tracks this way.

@@

During one week (five open clinic days), 894 patients were registered in the clinical database: four patients were bacteriologically diagnosed with TB and 16 patients were suspected of having TB after symptom screening. Overall, 1,563 person tracks were detected based on the video sensor data; 125 likely from clinic staff (movements from and to staff rooms) and 1,438 from clinic attendees (registered patients and other visitors). Most movements were detected in the waiting room (\Cref{fig:input-data-descriptives}a) and during the morning (\Cref{fig:input-data-descriptives}b). Attendees spent about half an hour in the clinic (median 25\,min, interquartile range (IQR) 13\,min$-$46\,min), mostly in the waiting room (\Cref{fig:input-data-descriptives}c). Close-contact encounters were frequent (median 5, IQR 2$-$9 per attendee) and attendees were in close contact with at least one other person for more than half of the duration of their visit (median 68\%, IQR 41\%$-$87\%). The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm (\supp~Fig.~\zref{fig:co2-levels}) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with a \eg median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room. Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).

>>>>

\title{\bfseries\LARGE\singlespacing{Estimating \emph{Mycobacterium} tuberculosis transmission in a South African clinic: Spatiotemporal model based on person movements}}
% author list
\author[1]{Nicolas Banholzer}
\author[2]{Keren Middelkoop}
\author[2]{Juane Leukes}
\author[3]{Ernest Weingartner}
\author[1]{Remo Schmutz}
\author[1]{Kathrin Zürcher}
\author[1,4,5]{Matthias Egger}
\author[2]{Robin Wood}
\author[1*]{Lukas Fenner}

@@

%TC:newcounter abst Words in abstract
%TC:envir abstract [] abst
\begin{abstract}\normalfont
The risk of Mycobacterium tuberculosis (\emph{Mtb}) transmission can be high in crowded clinics. We developed a spatiotemporal model of airborne \emph{Mtb} transmission, based on the Wells-Riley equation. We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when masks were compulsory and ventilation increased. We matched person movements with clinical records to identify the spatiotemporal location of infectious TB patients. We modeled the concentration of infectious doses (quanta) and estimated the individual risk of infection. Over five days, video sensors detected 1,438 clinic attendees. CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm); the quanta concentration was higher in the morning than in the afternoon, and highest in the waiting room. The estimated risk of infection per clinic attendee was 0.06\% (95\%-credible interval [CrI] 0.001\%$-$0.40\%). It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.28-fold (95\%-CrI 1.19$-$1.36) higher in scenarios without mask use and 1.74-fold (95\%-CrI 0.47$-$5.34) with pre-pandemic ventilation rates. Spatiotemporal modeling can identify high-risk areas and evaluate the impact of infection control measures in clinics.

@@

Based on a pilot study from 2019 in the same clinic\cite{Zurcher2022JID,Zurcher2020BMJ}, we collected environmental data (indoor CO$_2$ levels), clinical data (TB status), and video sensor data (person movements) for five days during the COVID-19 pandemic in October/November 2021 (October 13, 15, 25, and November 4, 5) at a primary care clinic in Cape Town, South Africa. The clinic offers both TB and HIV services and other basic clinical services, Monday through Friday, from 8\,am to 4\,pm. It is located in a large settlement of formal and semi-formal housing where TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}. We defined three areas within the clinic (\Cref{fig:floor-plan}): the waiting room (10.55m $\times$ 5.50m $\times$ 3.00m), corridor (12.45m $\times$ 2.20m $\times$ 2.50m), and TB room (4.75m $\times$ 3.50m $\times$ 3.00m). The TB room was mainly used for the treatment of diagnosed TB patients but also for the screening of suspected patients with respiratory symptoms who were ultimately not diagnosed with TB.

\begin{figure}[!htpb]
    \centering
    \includegraphics{doc/clinic-schematic-annotated-view.pdf}
    \caption{\textbf{Schematic view of the clinic}. Patients typically enter the clinic through the entrance on the bottom left, register at the reception, then wait in the waiting room or along the corridor until they see a doctor in one of the care rooms or the TB room. They exit the clinic through one of the exits in the waiting room or at the end of the corridor. Coloured icons show the placement of the video sensors to track person movements and the devices monitoring CO$_2$ levels.}
    \label{fig:floor-plan}
\end{figure}

@@

We used an anonymous personal tracking system using video sensors (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1-second intervals (\Cref{fig:floor-plan}). The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic. People in the clinic could contribute multiple tracks if they moved out of the range of a sensor or were not recognized as a person. To address this, we created an R Shiny application (\supp~Fig.~\zref{fig:shiny-app}) to rejoin interrupted tracks as described in \supp~Section~\zref{sec:setting-and-data}. We also labeled the tracks to identify the movements of clinic staff. \supp~Fig.~\ref{fig:tracking-examples} shows four examples of person movements after rejoining interrupted tracks.

\begin{figure}[!htpb]
    \centering
    \includegraphics{results/data/example-patient-tracks.png}
    \caption{Examples of person movements detected with our video sensors.}
    \label{fig:tracking-examples}
\end{figure}

@@

\begin{figure}[!htpb]
    \centering
    \includegraphics{doc/model-flow-chart.pdf}
    \caption{\textbf{Visual summary of the spatiotemporal modelling approach}. (1)~Clinical records and person movements were linked based on their timestamps at the registration to determine the spatiotemporal location of diagnosed (infectious) TB patients. (2)~Both diagnosed and randomly selected undiagnosed TB patients among clinic attendees generate infectious quanta. (3)~Person-tracking and environmental data were combined to estimate the air change rate. (4)~Air change rate was related to the speed at which quanta diffuses indoors. (5)~The air change rate also determines quanta removal through outdoor air exchange, in addition to inactivation of \emph{Mtb} in airborne particles. (6)~Quanta generation, diffusion, and removal determine the spatiotemporal quanta concentration, from which the personal risk of infection is calculated using the Wells-Riley equation.}
    \label{fig:modeling-flow}
\end{figure}

@@

(1)~We matched clinical records with person movements to identify the spatiotemporal location of infectious TB patients within the clinic. We recorded the timestamp when a person was detected in the registration area and matched this with the registration time in the clinical records. We excluded tracks with a total duration of less than five minutes in the clinic and a duration of less than five seconds in the registration area from the matching. We allowed for a maximum delay of 15\,min between the track's timestamp in the registration area and the entry in the clinical database. All diagnosed TB patients could be matched with person movements this way.

@@

During one week (five open clinic days), 894 patients were registered in the clinical database: four patients were bacteriologically diagnosed with TB and 16 patients were suspected of having TB after symptom screening. Overall, 1,563 person movements were detected based on the video sensor data; 125 likely from clinic staff (movements from and to staff rooms) and 1,438 from clinic attendees (registered patients and other visitors). Most movements were detected in the waiting room (\Cref{fig:input-data-descriptives}a) and during the morning (\Cref{fig:input-data-descriptives}b). Attendees spent about half an hour in the clinic (median 25\,min, interquartile range (IQR) 13\,min$-$46\,min), mostly in the waiting room (\Cref{fig:input-data-descriptives}c). Close-contact encounters were frequent (median 5, IQR 2$-$9 per attendee) and attendees were in close contact with at least one other person for more than half of the duration of their visit (median 68\%, IQR 41\%$-$87\%). The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm (\supp~Fig.~\zref{fig:co2-levels}) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with a \eg median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room. Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).
